---
title: "Validatetools"
subtitle: "Validatetools: Check and resolve contradictory rule sets" 
author: "Edwin de Jonge"
date: "Statistics Netherlands / eRum 2018"
output:
  beamer_presentation:
    includes:
      in_header: tex/header.tex
    keep_tex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
library(validatetools)
```

## Who am I?

- Data scientist / Methodologist Statistics Netherlands (aka CBS).
- Author of several R-packages, including `whisker`, `validate`, `errorlocate`, `docopt`, `tableplot`, `chunked`, `ffbase`,...
- Co-author of _Statistical Data Cleaning with applications in R (2018)_ (sorry for the plug, but relevant for this talk...)

## {.plain}

\hspace*{-1cm}\includegraphics[width=1\paperwidth]{img/bad-data}

## Data cleaning...

A large part of your and our job is spent in data-cleaning:

- getting your data in the right shape (e.g. `tidyverse`)

- checking validity (e.g. `validate`, `dataMaid`, `errorlocate`)

- impute values for missing or erroneous data (e.g. `VIM`, `simputation`)

- see data changes, improvements (e.g. `daff`, `lumberjack`)

### Desirable data cleaning properties: 

- Reproducible data checks
- Automate repeated data checking (e.g. monthly/quarterly stats)
- Monitor data improvements / changes. 
- But **How?**

## {.plain}

\hspace*{-1cm}
\includegraphics[width=1\paperwidth]{img/rules.png}

## Cleaning philosophy

- **"Explicit is better than implicit"**.
- Make data knowledge as **explicit** as possible.
- Store these as **validation rules**.

### Advantages:

- Easy checking of rules
- Data quality statistics: how often is each rule violated? 
- Allows for reasoning on rules: which variables are involved in errors?
- Simplifies rule changes and additions.

## Rules

- Data rules are solidified domain knowledge.
- In statistical production: much domain knowledge available.
- Many rules, many diverse subpopulations.
- Real world knowledge e.g. :

      - `age` is not negative.
      - human `age` is less then `150` years.
      
- Expert knowledge, e.g:

      - `IF profit > 0 THEN turnover > 0`
      - `IF married THEN age > 16` 


## R package `validate`

With package `validate` you formulate explicit rules that data must conform to:

```{r, eval=FALSE}
library(validate)
check_that( data.frame(age=160, job = "no", income = 3000), 
  age >= 0, 
  age < 150,
  job %in% c("yes", "no"),
  if (job == "yes") age >= 16,
  if (income > 0) job == "yes"
)
```

## Rules (2)

A lot of datacleaning packages are using validate rules to facilitate their work.

- `validate`: validation statistics
- `errorlocate` to find **errors** in variables (in stead of records)
- `rspa` data **correction** under data constraints
- `deductive` deductive **correction**
- `dcmodify` deterministic **correction** and **imputation**.

### Result: 

- Often many rules, great and small.
- Rules often defined multiple times at different processing steps.
- Rules may partially contradict each other.

## Why-o-why `validatetools`?

- We have package `validate`, what is the need?

### In "real life"

- We have large data analysis processes with many steps.
- Often many rules (> 100) on many variables (> 20) and many observations (> 0.5M).
- Results often in redundant or (partially) contradictive rule sets...

## Because we'd like to...

- clean up rule set ( kind of meta-cleaning...).
- detect and resolve problems with rules:

    - Detect **conflicting** rules.
    - Remove **redundant** rules.
    - **Substitute** values and **simplify** rules.
    - Detect unintended consequences.

- check the rule set using formal logic (without any data!).
- solve these kind of fun problems :-)

## Formally...

### Rule set $S$

A validation rule set $S$ is a conjunction of rules $r_i$, which applied on record $\la{x}$ returns `TRUE` (valid) or `FALSE` (invalid)

$$
S(\la{x}) = r_1(\la{x}) \land \cdots \land r_n(\la{x})
$$

### Note 

- a record has to comply to each rule $r_i$.

- it is thinkable that two or more $r_i$ are in conflict, making each record invalid.

## Formally...

### Rule $r_i(x)$

A rule a disjunction of atomic clauses:

$$
r_i(x) = \bigvee_j C_i^j(x)
$$
with: 

$$
C_i^j(\la{x}) = \left\{
  \begin{array}{l}
 \la{a}^T\la{x} \leq b \\
  \la{a}^T\la{x} = b \\
  x_j \in F_{ij}  \textrm{with } F_{ij} \subseteq D_j \\
  x_j \not\in F_{ij} \textrm{with } F_{ij} \subseteq D_j \\
  \end{array}
\right.
$$

## Mixed Integer Programming

Each problem we try to solve, can be translated into a mip problem, which can be 
readily solved using a mip solver. 

`validatetools` uses `lpSolveApi`.

$$
\begin{array}{r}
  \textrm{Minimize } f(\mathbf{x}) = 0; \\
  \textrm{s.t. }\mathbf{Rx} \leq \mathbf{d} \\
\end{array}
$$
with $\la{R}$ and $\la{d}$ the rule definitions and $f(\la{x})$ is the specific problem that is solved.

## Problem: infeasibility

### Problem

One or more rules in conflict:

- $S(x)$ always is `FALSE, **no data correct**
- _happens more often than you think_


```{r}
library(validatetools)
rules <- validator( is_adult = age >=21
                  , is_child = age < 18
                  )
is_infeasible(rules)
```

## {.plain}

\hspace*{-2cm}
\includegraphics[height=1\paperheight]{img/keepcalm-and-resolve.png}

## Conflict, and now?

```{r}
# Find out which rule would remove the conflict
detect_infeasible_rules(rules)
# And itś conflicting rule(s)
is_contradicted_by(rules, "is_adult")
```

- One of these rules needs to be removed
- Which one? Depends on human assessment...

## Detecting and removing redundant rules

- Rule $r_1$ may imply $r_2$, so $r_2$ can be removed.
- Simplifies rule set!

```{r}
rules <- validator( r1 = age >= 18
                  , r2 = age >= 12
                  )
detect_redundancy(rules)
remove_redundancy(rules)
```

## Value substitution

```{r}
rules <- validator( r1 = if (gender == "male") weight > 50
                  , r2 = gender %in% c("male", "female")
                  )

substitute_values(rules, gender = "male")
```

## Conditional statement

A bit more complex reasoning, but still classical logic:

```{r}
rules <- validator( r1 = if (income > 0) age >= 16
                  , r2 = age < 12
                  )
# age > 16 is always FALSE so r1 can be simplified
simplify_conditional(rules)
```

## All to gether now!:

`simplify_rules` applies all simplification methods to the rule set

```{r}
rules <- validator( r1 = if (age < 16) income == 0
                  , r2 = job %in% c("yes", "no")
                  , r3 = if (job == "yes") income > 0
                  )
simplify_rules(rules, job = "yes")
```

## Interested?

\begin{minipage}[c]{0.5\textwidth}
\includegraphics[width=0.9\textwidth]{img/SDCR.jpg} 
\end{minipage}
\begin{minipage}[c]{0.5\textwidth}
\begin{block}{SDCR}
M. van der Loo and E. de Jonge (2018)
\emph{Statistical Data Cleaning with applications in R}
Wiley, Inc.
\end{block}
\begin{block}{validatetools}
\begin{itemize}
\item Available on \href{https://CRAN.R-project.org/package=validatetools}{\underline{CRAN}}
\end{itemize}
\end{block}
\begin{block}{More theory?}
$\leftarrow$ See book 
\end{block}
\end{minipage}

Thank you for your attention! / Köszönöm a figyelmet!
